---
title: "Chapter 6 Applied Exercises"
author: "ZackBarry"
date: "7/6/2019"
output: 
  html_document:
  toc: true
toc_depth: 2
number_sections: true
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR)
library(dplyr)
library(tibble)
library(ggplot2)
library(MASS) # for lda()
library(class) # for knn()
library(boot) # for cv.glm() and boot()
library(MVN)  # for checking for multivariate normal distributions
# ^ need to install jags directly from 
#    https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/Mac%20OS%20X/
#   before installing MVN
library(leaps) # for regsubsets()
library(glmnet)  # for glmnet() (ridge/lasso regression)
library(pls)  # for pcr()

select <- dplyr::select # avoid namespace collision with MASS
filter <- dplyr::filter  
```

## 8) Best subset selection.

a) Use the `rnorm()` function to generate a predictor $X$ of length $n=100$, as
well as a noise vector $\epsilon$ of length $n=100$.

```{r}
set.seed(1)

X <- rnorm(100)
eps <- rnorm(100)
```

b) Generate a response vector $Y$ according to the model
$$ Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \epsilon $$
where $\beta_i$ are constants of your choice.

```{r}
Y = 4 + 1*X + -7*X^2 + 23*X^3 + eps
```

c) Use the `regsubsets()` funtion to perform best subset selection in order to 
choose the best model containing the predictors $X,X^2,\ldots,X^{10}$. What is
the best model obtained according to $C_p$, BIC, and adjusted $R^2$?
Show some plots to provide evidence for your answer, and report the coefficients
of the best model obtained. 

```{r}
all.data <- data.frame(X = X, Y = Y)
reg.fit <- regsubsets(Y ~ poly(X, degree = 10, raw = TRUE), all.data, nvmax = 10)

reg.sum <- summary(reg.fit)
print(paste(
  "The lowest BIC value was found when using", 
  which.min(reg.sum$bic),
  "variables."
))
print(paste(
  "The highest adjusted R2 value was found when using", 
  which.max(reg.sum$adjr2),
  "variables."
))
print(paste(
  "The lowest Cp value was found when using", 
  which.min(reg.sum$cp),
  "variables."
))
```
The adjusted training error statistics have different values indicate that the
inclusion of 3 or 4 variables results in the best fit. Let's look at some plots
to see their overall behavior:
```{r}
par(mfrow = c(1,3))
reg.plot <- data.frame(
  var.count = seq(1,10),
  bic = reg.sum$bic,
  adjr2 = reg.sum$adjr2,
  cp = reg.sum$cp
)
ggplot(reg.plot, aes(x = var.count, y = bic)) +
  geom_point() +
  geom_line()
ggplot(reg.plot, aes(x = var.count, y = adjr2)) +
  geom_point() +
  geom_line()
ggplot(reg.plot, aes(x = var.count, y = cp)) +
  geom_point() +
  geom_line()
par(mfrow = c(1,1))
```

In each of the plots, the difference between including 3 or 4 variables is
negligible.  We are tempted to use 3 variables to improve model interpretability 
and to avoid overfitting. Let's see what the coefficients of that model are:
```{r}
coef(reg.fit, 3)
```
The model is approximated as 
$$ Y = 4.06 + 0.98X - 7.12X^2 + 23.02 X^3 $$
which is very close to the true model. Let's consider the 4-variable fit as well:
```{r}
coef(reg.fit, 4)
```
The 4th order fit is also very close; it adds the 5th order term into the model, 
but uses a small coefficient.


d) Repeat (c) using forward and backward stepwise selection. Compare results.

```{r}
fwd.fit <- regsubsets(
  Y ~ poly(X, degree = 10, raw = TRUE), 
  all.data,
  nvmax = 10,
  method = "forward"
)
fwd.sum <- summary(fwd.fit)

bwd.fit <- regsubsets(
  Y ~ poly(X, degree = 10, raw = TRUE),
  all.data,
  nvmax = 10,
  method = "backward"
)
bwd.sum <- summary(bwd.fit)

print(paste("The min BIC for foward selection used",
            which.min(fwd.sum$bic),
            "variables; backward used",
            which.min(bwd.sum$bic)
           )
)
print(paste("The max adj. R2 for foward selection used",
            which.max(fwd.sum$adjr2),
            "variables; backward used",
            which.max(bwd.sum$adjr2)
           )
)
print(paste("The min Cp for foward selection used",
            which.min(fwd.sum$cp),
            "variables; backward used",
            which.min(bwd.sum$cp)
           )
)
```
The results for forward, backward, and best subset selection were the same in 
terms of the number of variables to minimize BIC and Cp and to maximize the
adjusted $R^2$ value. Let's compare the 3 variable model coefficients:

First, check that the variables used are the same:
```{r}
names(coef(fwd.fit, 3)) == names(coef(bwd.fit, 3))
names(coef(fwd.fit, 3)) == names(coef(reg.fit, 3))
```

Now we can construct a data frame of coefficient values with one variable per row:
```{r}
data.frame(
  subset.vars = names(coef(reg.fit, 3)),
  subset.sel = coef(reg.fit, 3),
  forward.sel = coef(fwd.fit , 3),
  backward.sel = coef(bwd.fit , 3)
)
```
Of course, all the coefficients agree since each of the fitting procedures is the
same; the coefficients would only differ if, say, forward selection used a quartic
term instead of a quadratic term.


e) Now fit a lasso model to the simulated data, again using $X$, $X^2$,..., $X^{10}$
as predictors. Use cross-validation to select the optimal value of $\lambda$.
Report the resulting coefficient estimates, and discuss the results obtained.

```{r}
cv.lambda <- cv.glmnet(
  x = poly(X, degree = 10, raw = TRUE),
  y = Y,
  alpha = 1
)

lasso.fit <- glmnet(
  x = poly(X, degree = 10, raw = TRUE),
  y = Y,
  alpha = 1
)

predict(lasso.fit, type = "coefficient", s = cv.lambda$lambda.min)
```
The lasso fit, which performs variable selection, successfuly selected the correct
variables.  However, the coefficient estimates were worse than the subset selection
method.

Note that in subset selection, the nonzero parameters will (only) be unbiased
if the removed predictors all have true coefficient values of 0. Lasso
regression has bias introduced due to the selection of the parameter $\lambda$.
In the example above, where subset selection resulted in unbiased coefficients
it is not suprising that it outperformed lasso regression.  However, if we
did not know the true model, it would be much hard to tell if this were the case.
It would probably be better to stick with lasso regression since it errs on the
side of including more potential predictors (depending on how large $\lambda$ is).

f) Now generate a response vector $Y$ according to the model
$$ Y =\beta_0+\beta_7 X^7 + \epsilon$$
and perform best subset selection and the lasso. Discuss the results.

```{r}
Y = 4 + 33 * X^7 + eps
```

We'll do the best subset method first:
```{r}
reg.fit <- regsubsets(
  Y ~ poly(X, power = 10, raw = TRUE),
  data.frame(Y = Y, X = X),
  nvmax = 10
)
reg.sum <- summary(reg.fit)
which.min(reg.sum$bic)
which.min(reg.sum$cp)
which.max(reg.sum$adjr2)
```
The three adjusted training errors disagree with regards to the number of variables
to include in the model. Let's look at their graphs to see if we can find a
good comprimise:
```{r}
errs <- data.frame(
  var.count = seq(1, 10),
  bic = reg.sum$bic,
  cp = reg.sum$cp,
  adjr2 = reg.sum$adjr2
)
par(mfrow = c(1,3))
ggplot(errs, aes(x = var.count, y = bic)) +
  geom_point() +
  geom_line()
ggplot(errs, aes(x = var.count, y = cp)) +
  geom_point() +
  geom_line()
ggplot(errs, aes(x = var.count, y = adjr2)) +
  geom_point() +
  geom_line()
par(mfrow = c(1,1))
```
Adjusted $R^2$ performs best at $N=3,4$, but BIC steadily increases as $N$ increases;
$C_p$ is similarly small for $N=1,2,3$. We select $N=2$ as a comprimise. Let's
look at the coefficients for that model:
```{r}
coef(reg.fit, 2)
```
This is very close to the true model. Let's see how it compares to lasso regression
with $\lambda$ selected via cross validation:
```{r}
cv.lambda <- cv.glmnet(
  x = poly(X, power = 10, raw = TRUE),
  y = Y,
  alpha = 1
)
lasso.fit <- glmnet(
  x = poly(X, power = 10, raw = TRUE),
  y = Y,
  lambda = cv.lambda$lambda.min
)
predict(lasso.fit, type = "coefficient", s = cv.lambda$lambda.min)
```
Lasso regression performs slightly worse than best subset selection for the same
number of coefficients. See the discussion at the end of part (e).

We could have reasonably selected $N=3$ for best subset selection, though, lets
see how that performs: 
```{r}
coef(reg.fit, 3)
```
It still outperforms lasso regression.


## PCR using `ISLR::College`

We will be predicting the number of applications received.

a) Split the data into a training set and a test set.
```{r}
train <- sample(c(TRUE, FALSE), nrow(College), replace = T, prob = c(0.8, 0.2))

train.set <- College[train, ]
test.set <- College[!train, ]
```

b) Fit a linear model using least squares on the training set, and report the test
error obtained.
```{r}
linear.fit <- lm(Apps ~ ., data = train.set)
summary(linear.fit)
test.pred <- predict(linear.fit, test.set)
linear.mse <- mean((test.pred - test.set$Apps)^2)
print(paste("The test MSE for the linear model is", linear.mse))
```
The test MSE is very poor though most of the coefficients are statistically
significant.


c) Fit a ridge regression model on the training set, with $\lambda$ chosen
by cross-validation. Report the test error obtained.

Recall that to use `glmnet()`, we need to create a matrix version of $X$ in which the 
categorical predictors are already coded to dummy variables.
```{r}
train.mat <- model.matrix(Apps ~ ., train.set)[, -1]

lambda.cv <- cv.glmnet(
  x = train.mat,
  y = train.set$Apps,
  alpha = 0
)
ridge.fit <- glmnet(
  x = train.mat,
  y = train.set$Apps,
  alpha = 0
)
ridge.coeff <- predict(ridge.fit, type = "coefficient", s = lambda.cv$lambda.min)
ridge.mse <- mean((model.matrix(Apps ~ ., test.set) %*% ridge.coeff - test.set$Apps)^2)
ridge.mse
```
The test MSE for ridge regression is slightly higher than the ordinary least
squares solution.


d) Fit a lasso model on the training set, with $\lambda$ chosen by cross-validation.
Report the test error obtained, along with the number of non-zero coefficient
estimates.

All we need to do is repeat the process for ridge regression with `alpha = 1` 
instead of `alpha = 0`. Report the test error obtained, along with the number
of non-zero coefficient estimates.
```{r}
train.mat <- model.matrix(Apps ~ ., train.set)[, -1]

lambda.cv <- cv.glmnet(
  x = train.mat,
  y = train.set$Apps,
  alpha = 1
)
lasso.fit <- glmnet(
  x = train.mat,
  y = train.set$Apps,
  alpha = 1
)
lasso.coeff <- predict(lasso.fit, type = "coefficient", s = lambda.cv$lambda.min)
lasso.mse <- mean((model.matrix(Apps ~ ., test.set) %*% lasso.coeff - test.set$Apps)^2)
lasso.mse
```
The test MSE for the lasso fit is an improvement over both OLS and ridge 
regression. Let's see how many coefficients it set to zero:
```{r}
lasso.coeff
```
Lasso regression set `F.Undergrad`, `Books`, and `perc.alumni` to zero.
We note that neither `Books` nor `perc.alumni` had statistically significant
coefficients in the OLS fit.


e) Fit a PCR model on the training set, with $M$ chosen by cross-validation.
Report the test error obtained, along with the value of $M$ as selected by 
cross-validation.

Recall that we need to center the variables before applying Primary Component
Regression.
```{r}
pcr.fit <- pcr(
  Apps ~ ., 
  data = train.set, 
  validation = "CV",
  center = TRUE
)
summary(pcr.fit)
```
95% of the variance in the predictors is explained by using just 3 components, 
and 100% is explained using 9 components. Similarly, the percent variance 
of `Apps` begins to level off after 5 components are used. Let's consider the
behavior of the mean squared error:
```{r}
validationplot(pcr.fit, val.type = "MSEP")
```
MSE starts to level off after 6 components are added. Let's calculate the
test error:
```{r}
pcr.pred <- predict(pcr.fit, test.set, ncomp = 6)
pcr.mse <- mean((pcr.pred - test.set$Apps)^2)
pcr.mse
```
PCR performs worse than ridge and lasso regression and OLS regression.


f) Fit a PLS model on the training set, with $M$ chosen by cross-validation.
Report the test error obtained, along with the value of $M$ selected by
cross-validation.

```{r}
pls.fit <- plsr(
  Apps ~ ., 
  data = train.set, 
  validation = "CV",
  center = TRUE
)
summary(pls.fit)
```
PLS has similar variance results for the training set as in PCR.

```{r}
validationplot(pls.fit, val.type = "MSEP")
```
The training MSE for PLS levels off a bit earlier than for PCR, with $M=5$ 
seeming to be a good fit.

```{r}
pls.pred <- predict(pls.fit, test.set, ncom = 5)
pls.mse <- mean((pls.pred - test.set$Apps)^2)
pls.mse
```
PLS performs slightly worse than PCR.


g) Comment on the results obtained. How accurately can we predict the number of
college applications received? Is there much difference among the test errors
resulting from these five approaches?

First, let's get a picture of how close the results of the best fitting model 
(that which used lasso regression) compare to the true values.
```{r}
lasso.pred <- as.vector(model.matrix(Apps ~ ., test.set) %*% lasso.coeff)
ggplot(data.frame(x = test.set$Apps, y = lasso.pred), aes(x, y)) +
  geom_point() +
  labs(title = "lasso regression results vs true Apps values",
       x = "true value",
       y = "lasso regression"
  )
ggplot(data.frame(x = test.set$Apps, y = lasso.pred - test.set$Apps), aes(x, y)) +
  geom_point() +
  labs(title = "lasso regression residuals vs true Apps values",
       x = "true value",
       y = "residuals"
  )
```
The residuals are slightly heteroscedastic, but uncorrelated with mean close to zero.
The lasso model appears to be a good fit for colleges where `Apps` is larger than
2500, but poor for colleges where `Apps` is near zero.


## Test vs. Training Error as features grow

We have seen that as the number of features used in a model increases, the training 
error will necessarily decrease, but the test error may not. We will now explore
this in a simulated data set.

a) Generate a data set with $p=20$ features, $n=1000$ observations, and an
associated quantitative response vector generated according to the model 
$$ Y = X\beta + \epsilon$$
where $\beta$ has some elements that are exactly equal to zero.

```{r}
set.seed(7)

p = 20
n = 1000

mydata <- matrix(rnorm(n * p), n, p)
colnames(mydata) <- paste("x", seq(1, 20), sep = "")
mydata <- as.data.frame(mydata) %>%
  mutate(x2 = sqrt(abs(x2) + 0.001),
         x3 = x3^2,
         x4 = x4^3 / 200,
         x7 = x7 * x2 / 10,
         x9 = abs(x9) ^ (1/4),
         x11 = x11 ^ (-1),
         x12 = x9 * x11,
         x13 = x2 / (x3 + 0.1),
         x17 = rnorm(n, mean = 4, sd = 6),
         x18 = x17 + runif(1000, -0.1, 0.1),
         x19 = 3 * x3 + runif(1000, 4.5, 4.6)) %>%
  as.matrix()

B <- rnorm(p)
B[c(1, 5, 6, 8, 10, 14, 17)] <- 0
y <- mydata %*% B + rnorm(p)

mydata <- as.data.frame(mydata) %>%
  mutate(y = as.vector(y))
```

b) Split your data set into a training set containing 100 observations and a test
set containing 900 observations.

```{r}
set <- sample(c("test", "train"), 1000, replace = T, prob = c(0.1, 0.9))
train <- mydata[set == "train", ]
test <- mydata[set == "test", ]
```

c) Perform best subset selection on the training set, and plot the training and
test set MSE associated with the best model of each size.

```{r}
# Apply best subset selection.
subset.fit <- regsubsets(
  y ~ .,
  data = train,
  nvmax = 20
)

# Create a matrix of coefficients.
coef <- matrix(0, nrow = 20, ncol = 21)
var_names <- names(coef(subset.fit, 20))
for (i in 1:20) {
  coef_i <- coef(subset.fit, i)
  coef[i, var_names %in% names(coef_i)] <- coef_i
}
colnames(coef) <- var_names

# Turn the train and test sets into matrices, leading with a column of 1s
# for the intercept term.
train_mat <- matrix(c(rep(1, dim(train)[1]), as.matrix(select(train, -y))),
                nrow = dim(train)[1], ncol = 21)
test_mat <- matrix(c(rep(1, dim(test)[1]), as.matrix(select(test, -y))),
                nrow = dim(test)[1], ncol = 21)

# Estimate y for the training and test sets.
train_est_y <- train_mat %*% t(coef)
test_est_y <- test_mat %*% t(coef)

# Calculate the MSE.
train_mse <- apply(train_est_y, 2, function(est) { mean((est - train$y)^2) })
test_mse <- apply(test_est_y, 2, function(est) { mean((est - test$y)^2) })

# Plot the MSE.
par(mfrow = c(1, 2))
p <- data.frame(
    train_mse = train_mse,
    test_mse = test_mse,
    feature_count = seq(1, 20)
  ) %>%
    ggplot(aes(x = feature_count)) 
p +
  geom_line(aes(y = train_mse)) +
  geom_point(aes(y = train_mse)) +
  labs(title = "train MSE")
p + 
  geom_line(aes(y = test_mse)) +
  geom_point(aes(y = test_mse)) +
  labs(title = "test MSE")
par(mfrow = c(1, 1))
```



e) For which model size does the test set MSE take on its minimum value? 
Comment on your results. If it takes on its minimum value for a model containing
only an intercept or a model containing all of the features, then play around
with the way you are generating the data in (a) until you come up with a scenario
in which the test set MSE is minimized for an intermediate model size.

```{r}
which.min(test_mse)
```
The minimum test MSE is lower than the number of true nonzero variables (13), but
seems to take into account that there are 3 pairs of highly correlated variables
(so selecting one of each pair leads to a better result than keeping all 6).

f) How does the model at which the test set MSE is minimized compare to the true
model used to generate the data? Comment on the coefficient values.

```{r}
coef(subset.fit, which.min(test_mse))
names(B) <- names(coef(subset.fit, 20))[-1]
B
```
The coefficient values for the variables included in the 9-feature best subset
model are very close to the true corresponding coefficient values.  Notably,
`x3`, `x4`, `x7`, `x9` were all set to zero by the best subset model, but are 
nonzero in the true model.  This makes sense for `x3` because it is constructed
to be highly correlated with `x19`:
```{r}
cor(mydata$x3, mydata$x19)
```



mydata <- as.data.frame(mydata) %>%
  mutate(x2 = sqrt(abs(x2) + 0.001),
         x3 = x3^2,
         x4 = x4^3 / 200,
         x7 = x7 * x2 / 10,
         x9 = abs(x9) ^ (1/4),
         x11 = x11 ^ (-1),
         x12 = x9 * x11,
         x13 = x2 / (x3 + 0.1),
         x17 = rnorm(n, mean = 4, sd = 6),
         x18 = x17 + runif(1000, -0.1, 0.1),
         x19 = 3 * x3 + runif(1000, 4.5, 4.6)) %>%
  as.matrix()


g) Create a plot displaying $\sqrt{\sum_{j=1}^p(\beta_j-\hat{\beta}_j^r)^2}$
for a range of values of $r$, where $\hat{\beta}_j^r$ is the $j$th coefficient
estimate for the best model containing $r$ coefficients. Comment on what you 
observe. How does this compare to the test MSE plot from (d)?









